High-order cognition is supported by complex but compressible brain activity patterns

### 2023-02-27 paper/boneyard.tex
finished integrating figures, including drafts of all captions.  also did some basic cleanup.

@@

>>>>

- We trained classifiers using more and more principle components to decode, and compared across condi- tions with varying degrees of cognitive richness.
-We found that as listening conditions become more cognitively rich,
decoding accuracy increased.
-Also, decoding accuracy increased as understanding of the narrative
accumulated over time, in more complex listening conditions.
- Decoding accuracy also increased in higher cognitive areas, in more complex listening conditions.
-We found that as story listening conditions become more complex, more
components are required to decode.
-We also found we could decode better with more impoverished data when
there is the underlying structure of the narrative providing more cognitive richness.
-We posit that as the complexity of our thoughts increases, neural
compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

Based on prior work ~\citep{Deme19} and following the direction of the field ~\citep{Turk13} we think our thoughts might be encoded in
dynamic network patterns, and possibly higher order network
patterns (Fig.~\ref{fig:direction_of_field}). We sought to test this hypothesis by developing an approach
to inferring high-order network dynamics from timeseries data.

One challenge in studying dynamic interactions is the
computational resources required to calculate higher-order correlations. 
We developed a computationally tractable model of network dynamics (Fig.~\ref{fig:methods_fig}) that takes in a feature
timeseries and outputs approximated first-order dynamics (i.e.,
dynamic functional correlations), second-order dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).

We first validated our model using synthetic data, and explored how
recovery varied with different underlying data structures and kernels.   We then 
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio
recording of a story, as well as scrambled versions of the same story
(where the scrambling was applied at different temporal scales).  We
trained classifiers to take the output of the model and decode the
timepoint in the story (or scrambled story) that the participants were
listening to. We found that, during the intact listening condition in the
experiment, classifiers that incorporated higher-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  a.\&d.).  By contrast, these
higher-order correlations were not necessary to support decoding the other
listening conditions and (minimally
above chance) during a control rest condition.  This suggests
that the cognitive processing that supported the most cogntively rich listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best when incorporating
higher-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique.  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions. Exploring this method
  with other data collection technique will be important to
  disentangle this question.

  \subsection*{Concluding remarks}

How can we better understand how brain patterns change over
time? How can we quantify the potential network dynamics that might be
driving these changes? One way to judge the techniques of the future is
to look at the trajectory of the fMRI field so far has taken so far
(Fig.~\ref{fig:methods_fig}).  The field started with 
univariate activation, measuring the average activity for each voxel.
Analyses of multivariate activation followed, looking at spatial patterns of
activity over voxels. Next, correlations of activity were explored, first
with measures like resting connectivity that take temporal correlation
between a seed voxel and all other voxels then with full connectivty
that measure all pairwise correlations.  Additionally, this path of increasing
complexity also moved from static to dynamic measurements.  One
logical next step in this trajectory would be dynamic higher-order
correlations. We have created a method 
to support these calculations by scalably approximating dynamic higher-order
correlations.

@@

We were also curious how compression would change across time.  If, there is some understanding of the narrative that accumulates over time, we should be able to see that difference. We found increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
We also found decreases in decoding accuracy for the word-scrambled and rest
condition.

Overall, we found that as story listening conditions become more complex, more components are required to decode. We also found we could decode better with more impoverished data when
there is the underlying structure of the narrative providing more cognitive richness. We posit that as the complexity of our thoughts increases, neural compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

% - Increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
% - Decreases in decoding accuracy for the word-scrambled and rest
% condition.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-02-27 paper/main.tex
finished integrating figures, including drafts of all captions.  also did some basic cleanup.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{figs/information_and_compression_over_time.pdf}

@@

We were also curious how compression would change across time.  If, there is some understanding of the narrative that accumulates over time, we should be able to see that difference. We found increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
We also found decreases in decoding accuracy for the word-scrambled and rest
condition.

Overall, we found that as story listening conditions become more complex, more components are required to decode. We also found we could decode better with more impoverished data when
there is the underlying structure of the narrative providing more cognitive richness. We posit that as the complexity of our thoughts increases, neural compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

% - Increases in decoding accuracy with the same number or fewer components for more complex, cognitively rich, conditions.
% - Decreases in decoding accuracy for the word-scrambled and rest
% condition.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

- We trained classifiers using more and more principle components to decode, and compared across condi- tions with varying degrees of cognitive richness.
-We found that as listening conditions become more cognitively rich,
decoding accuracy increased.
-Also, decoding accuracy increased as understanding of the narrative
accumulated over time, in more complex listening conditions.
- Decoding accuracy also increased in higher cognitive areas, in more complex listening conditions.
-We found that as story listening conditions become more complex, more
components are required to decode.
-We also found we could decode better with more impoverished data when
there is the underlying structure of the narrative providing more cognitive richness.
-We posit that as the complexity of our thoughts increases, neural
compression decreases. However, as our thoughts become deeper and richer, more reliable information is available at higher neural compression.

Based on prior work ~\citep{Deme19} and following the direction of the field ~\citep{Turk13} we think our thoughts might be encoded in
dynamic network patterns, and possibly higher order network
patterns (Fig.~\ref{fig:direction_of_field}). We sought to test this hypothesis by developing an approach
to inferring high-order network dynamics from timeseries data.

One challenge in studying dynamic interactions is the
computational resources required to calculate higher-order correlations. 
We developed a computationally tractable model of network dynamics (Fig.~\ref{fig:methods_fig}) that takes in a feature
timeseries and outputs approximated first-order dynamics (i.e.,
dynamic functional correlations), second-order dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).

We first validated our model using synthetic data, and explored how
recovery varied with different underlying data structures and kernels.   We then 
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio
recording of a story, as well as scrambled versions of the same story
(where the scrambling was applied at different temporal scales).  We
trained classifiers to take the output of the model and decode the
timepoint in the story (or scrambled story) that the participants were
listening to. We found that, during the intact listening condition in the
experiment, classifiers that incorporated higher-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  a.\&d.).  By contrast, these
higher-order correlations were not necessary to support decoding the other
listening conditions and (minimally
above chance) during a control rest condition.  This suggests
that the cognitive processing that supported the most cogntively rich listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best when incorporating
higher-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique.  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions. Exploring this method
  with other data collection technique will be important to
  disentangle this question.

  \subsection*{Concluding remarks}

How can we better understand how brain patterns change over
time? How can we quantify the potential network dynamics that might be
driving these changes? One way to judge the techniques of the future is
to look at the trajectory of the fMRI field so far has taken so far
(Fig.~\ref{fig:methods_fig}).  The field started with 
univariate activation, measuring the average activity for each voxel.
Analyses of multivariate activation followed, looking at spatial patterns of
activity over voxels. Next, correlations of activity were explored, first
with measures like resting connectivity that take temporal correlation
between a seed voxel and all other voxels then with full connectivty
that measure all pairwise correlations.  Additionally, this path of increasing
complexity also moved from static to dynamic measurements.  One
logical next step in this trajectory would be dynamic higher-order
correlations. We have created a method 
to support these calculations by scalably approximating dynamic higher-order
correlations.

\section*{Acknowledgements}
We acknowledge discussions with Luke Chang, Hany Farid, Paxton
Fitzpatrick, Andrew Heusser, Eshin Jolly, Qiang Liu, Matthijs van der
Meer, Judith Mildner, Gina Notaro, Stephen Satterthwaite, Emily
Whitaker, Weizhen Xie, and Kirsten Ziman. Our work was supported in
part by NSF EPSCoR Award Number 1632738 to J.R.M. and by a sub-award
of DARPA RAM Cooperative Agreement N66001-14-2-4-032 to J.R.M.  The
content is solely the responsibility of the authors and does not
necessarily represent the official views of our supporting
organizations.

\section*{Author contributions}
Concept: J.R.M. and L.L.W.O. Implementation: L.L.W.O., and
J.R.M.  Analyses: L.L.W.O and J.R.M.

>>>>

\newcommand{\synthThirds}{S1}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{figs/information_and_compression_over_time}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pca_neurosynth}

\caption{\textbf{Top terms associated with the highest-weighted components by
condition.} Each row corresponds to an experimental condition, and the colors
correspond to the component number (ranked by proportion of variance
explained). The inflated brain plots display the spatial distributions of each
components' hubs (see \textit{Topographic Factor Analysis}). The lists on the
right display the top five Neurosynth terms~\citep{RubiEtal17} decoded from
each components' brain map. Analogous maps computed separately for each story
segment may be found in Figure~\synthThirds.}

\label{fig:neurosynth-pca}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/pca_changes_neurosynth}

\caption{\textbf{Changes in decoding accuracy and compression over time.} Each
row corresponds to an experimental condition, and the colors correspond to
change direction (positive: red; negative: blue). The saturation denotes the
component number (columns on the right).  The lists on the
right display the top five Neurosynth terms~\citep{RubiEtal17} decoded from
each components' brain map.}

\label{fig:neurosynth-changes}

\section*{Discussion}

\section*{Methods}

\section*{Acknowledgements}

We acknowledge discussions with Rick Betzel, Emily Finn, and Jim Haxby. Our
work was supported in part by NSF CAREER Award Number 2145172 to J.R.M. The
content is solely the responsibility of the authors and does not necessarily
represent the official views of our supporting organizations. The funders had
no role in study design, data collection and analysis, decision to publish, or
preparation of the manuscript.

\section*{Author contributions}

Conceptulization: J.R.M. and L.L.W.O. Methodology: J.R.M. and L.L.W.O.
Implementation: L.L.W.O. Analysis: L.L.W.O. Writing, Reviewing, and Editing:
J.R.M. and L.L.W.O. Supervision: J.R.M.
